---
title: "5301-Final Report"
author: "Soham Neeraj Agarkar (1002157894) & Utkarsh Pant (1002170893"
date: "2023-12-12"
output: 
  pdf_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Effect of Gender and Diet on Weight Loss

## General Declaration

```{r}
library(WRS2)
library(patchwork)
library(car)
library(outliers)
library(dplyr)
library(ggplot2)
```

\newpage

## Introduction

In today's world there is a very heavy focus on weight loss. Scores of corporations are constantly locked in a struggle to make products that target factors such as weight gain, weight loss and general beauty products and market them to a wide spectrum of audiences. In such a market, data analysis could play a crucial role and allow corporations to make an informed decision on what their product needs versus what their customers want.

Today we set out to understand which attributes are the most likely to affect a person's weight loss. We will work with the diet data set that is already a present in the [WRS2] package.

## Problem Statement

Perform Two-Way ANOVA on the features *gender* and *diet.type* to determine which is more influential to *weight.loss* feature.

\newpage

## Descriptive Analysis

Below is the distribution of our data's descriptive analysis :

```{r}
str(diet)
```

We can observe from the above the analysis the structure of our data set. The data possesses two categorical variables and the rest are numerical.

```{r}
summary(diet$gender)
```

```{r}
summary(diet$age)
```

```{r}
summary(diet$height)
```

```{r}
summary(diet$diet.type)
```

```{r}
summary(diet$initial.weight)
```

```{r}
summary(diet$final.weight)
```

```{r}
summary(diet$weight.loss)
```

Here we see the feature wise summary statistics.

Below we see the features statistics while being grouped by ***gender** :*

```{r}
diet %>%
  group_by(diet$gender) %>%
  summarise(
    Mean = mean(diet$age),
    Median = median(diet$age),
    SD = sd(diet$age),
    Min = min(diet$age),
    Max = max(diet$age)
  )
```

```{r}
diet %>%
  group_by(diet$gender) %>%
  summarise(
    Mean = mean(diet$height),
    Median = median(diet$height),
    SD = sd(diet$height),
    Min = min(diet$height),
    Max = max(diet$height)
  )
```

```{r}
diet %>%
  group_by(diet$gender) %>%
  summarise(
    Mean = mean(diet$initial.weight),
    Median = median(diet$initial.weight),
    SD = sd(diet$initial.weight),
    Min = min(diet$initial.weight),
    Max = max(diet$initial.weight)
  )
```

```{r}
diet %>%
  group_by(diet$gender) %>%
  summarise(
    Mean = mean(diet$final.weight),
    Median = median(diet$final.weight),
    SD = sd(diet$final.weight),
    Min = min(diet$final.weight),
    Max = max(diet$final.weight)
  )
```

```{r}
diet %>%
  group_by(diet$gender) %>%
  summarise(
    Mean = mean(diet$weight.loss),
    Median = median(diet$weight.loss),
    SD = sd(diet$weight.loss),
    Min = min(diet$weight.loss),
    Max = max(diet$weight.loss)
  )
```

Below we see the feature statistics while grouped by ***diet.type*** :

```{r}
diet %>%
  group_by(diet$diet.type) %>%
  summarise(
    Mean = mean(diet$age),
    Median = median(diet$age),
    SD = sd(diet$age),
    Min = min(diet$age),
    Max = max(diet$age)
  )
```

```{r}
diet %>%
  group_by(diet$diet.type) %>%
  summarise(
    Mean = mean(diet$height),
    Median = median(diet$height),
    SD = sd(diet$height),
    Min = min(diet$height),
    Max = max(diet$height)
  )
```

```{r}
diet %>%
  group_by(diet$diet.type) %>%
  summarise(
    Mean = mean(diet$initial.weight),
    Median = median(diet$initial.weight),
    SD = sd(diet$initial.weight),
    Min = min(diet$initial.weight),
    Max = max(diet$initial.weight)
  )
```

```{r}
diet %>%
  group_by(diet$diet.type) %>%
  summarise(
    Mean = mean(diet$final.weight),
    Median = median(diet$final.weight),
    SD = sd(diet$final.weight),
    Min = min(diet$final.weight),
    Max = max(diet$final.weight)
  )
```

```{r}
diet %>%
  group_by(diet$diet.type) %>%
  summarise(
    Mean = mean(diet$weight.loss),
    Median = median(diet$weight.loss),
    SD = sd(diet$weight.loss),
    Min = min(diet$weight.loss),
    Max = max(diet$weight.loss)
  )
```

**Frequency of Categorical variables** :

```{r}
# Frequency table for a gender
table(diet$gender)

# Percentage table for a gender
prop.table(table(diet$gender)) * 100

```

```{r}
# Frequency table for a diet.type
table(diet$diet.type)

# Percentage table for a categorical variable
prop.table(table(diet$diet.type)) * 100

```

**Checking for missing values :**

```{r}
colSums(is.na(diet))
```

As evidenced by the result of the code above, there are **no missing values** in our data set.

\newpage

## Plots and Visualizations

**Plots for Gender**

```{r}
#specifying colors for the plot
gender_colors <- c("hotpink", "blue")

#histogram weightloss by gender 
p1 <- ggplot(data = diet, aes(x = weight.loss, fill = gender)) +
  geom_histogram(position = "identity", alpha = 0.65) +
  labs(title = "Weight Loss vs Gender", x = "Weight Loss", y = "Count") +
  scale_fill_manual(values = gender_colors) +
  theme_minimal() 

#density plot for weightloss by diet.type
p2 <- ggplot(data = diet, aes(x=weight.loss, fill = gender)) +
  geom_density(aes(x=weight.loss, color = gender), position  = "identity", alpha = 0.65, linewidth = 1) +
  labs(title = "Weight Loss vs Gender", x= "Weight Loss") +
  scale_fill_manual(values = gender_colors) +
  theme_minimal()

#Gender plots
p1 + p2

```

**Box plot for Gender**:

```{r}
#boxplot for weightloss by gender
ggplot(data =diet, aes(x=weight.loss, fill = gender)) +
  geom_boxplot(aes(x = gender, y=weight.loss, group = gender)) +
  labs(title = "Weight Loss grouped by Gender", x="Gender", y = "Weight Loss") + 
  scale_fill_manual(values = gender_colors) +
  theme_minimal()
```

**Plots for Diet Type:**

```{r}
#diet colors
diet_colors <- c("#b30000", "#0d88e6", "#ebdc78")

#Histogram plots for weight loss by diet type
p3 <- ggplot(data= diet, aes(x = diet$weight.loss, fill = diet$diet.type)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  labs(title = "Weight Loss by Diet Type", x="Weight Loss", y="Count", fill = "Diet Type") +
  scale_fill_manual(values = diet_colors) +
  theme_minimal()

p4 <- ggplot(data = diet, aes(x = diet$weight.loss, fill = diet$diet.type)) + 
  geom_density(aes(x=weight.loss), position = "identity", alpha =0.65, linewidth =1) +
  labs(title = "Weight Loss by Diet Type", x= "Weight Loss", fill = "Diet Type") +
  scale_fill_manual(values = diet_colors) +
  theme_minimal()

#Plots for weight loss by diet type
p3 + p4
```

**Box plot for Diet Type:**

```{r}
#boxplot for weight loss by diet type
ggplot(data = diet, aes(x=weight.los, fill = diet.type)) + 
  geom_boxplot(aes(x=diet.type, y=weight.loss, group = diet.type)) +
  labs(title = "Boxplot for Weightloss by Diet Type", x = "Diet Type", y = "Weight Loss", fill = "Diet Type") + 
  scale_fill_manual(values = diet_colors) +
  theme_minimal()
```

The above box plot for the *diet.type* feature shows some outliers in diet type 'A' and 'B'. Simply eyeballing it we can assume that while 'A' appears to have outliers, it is important to keep in mind that those values are only outliers among instances that follow diet type 'A'. If we consider the whole feature, those values aren't considered outliers as the values for diet type 'C' stretches all the way up to include those values.

Diet type 'B', on the other hand, seems to have one outlier that falls far below the lowest value in the entire feature among all categories ('A', 'B', 'C').

**Outlier Detection:**

Now that we know that the feature has one outlier, we can try to determine what outlier it is. Since we only have to detect one outlier, we can use the grubbs single outlier detection test.

```{r}
#outlier detection
grubbs.test(diet$weight.loss)
```

The above code displays the outlier in the data. Since there is only outlier in the data, we chose to leave it in as it doesn't affect the overall outcome of the analysis.

\newpage

## ANOVA

ANOVA, or Analysis of Variance, is a statistical method used to analyze the differences among group means in a sample. It is an extension of the t-test, allowing for the comparison of means across more than two groups. ANOVA assesses whether there are any statistically significant differences between the means of three or more independent (unrelated) groups.

### Null and Alternate Hypotheses

**Null Hypothesis (**$H_O$**)** : Means of groups by *gender* are similar, means of groups by *diet.type* are similar and there is no significant interaction between *gender* and *diet.type* features.

**Alternate Hypothesis (**$H_a$**)** : There is a significant difference betweeen the means of atleast one factor or there is significant interaction between *gender* and *diet.type.*

### Assumptions for ANOVA

ANOVA makes the following assumptions :

Independance of of observations

:   Since we know that the data has been collected at random and without any bias we can surmise that that the observations are independent.

Normality

:   ANOVA assumes that the data within each group are normally distributed.

Homogeneity of Variance

:   This means that the variability within each group is roughly the same.

### Pre-Analysis Testing for Weight Loss grouped by Gender

```{r}
#Normality Test for Gender
shapiro.test(diet$weight.loss[diet$gender == "Female"])
shapiro.test(diet$weight.loss[diet$gender == "Male"])
```

If we observe the p-values acquired from the test, we can conclude that even when grouped the data is still somewhat normally distributed. We say this because our $\alpha$ = 0.05 and both p-values are greater than 0.05.

```{r}
#Levene's Test for Homogeneity of Variance
leveneTest(data = diet, group = diet$gender, y= diet$weight.loss)
```

Since, our p-value is greater than $\alpha$ (0.05), we can conclude that there exists Homogeneity of Variances in the *weight.loss* feature when grouped by *gender*.

In the case of *weight.loss* grouped by *gender*, all our tests have yielded posisitve.

### Pre-Analysis Testing for Weight Loss grouped by Diet Type

```{r}
#Normality Test for Diet Type
shapiro.test(diet$weight.loss[diet$diet.type == "A"])
shapiro.test(diet$weight.loss[diet$diet.type == "B"])
shapiro.test(diet$weight.loss[diet$diet.type == "C"])

```

In all cases, the p-value is greater than $\alpha$ (0.05). Hence we can say that the data is normally distributed even when split into groups.

```{r}
#Levene's Test for Homogeneity of Variance
leveneTest(data = diet, group = diet$diet.type, y= diet$weight.loss)
```

Since, our p-value is greater than $\alpha$ (0.05), we can conclude that there exists Homogeneity of Variances in the *weight.loss* feature when grouped by *diet.type*.

In the case of *weight.loss* grouped by *diet.type,* all tests have yielded positive.

Having met all assumptions, we can now move forward with main ANOVA test itself.

### ANOVA Test

```{r}
#Two-way ANOVA
model <- aov(weight.loss ~ gender * diet.type, data=diet)

summary(model)
```

Judging by the results of the ANOVA, we can conclude that ***diet.type*** plays an important role in the analysis. ***gender*** by itself has no importance but when taken in conjunction with ***diet.type*** it may be a facotr in weight loss.

Thus, we reject the null hypothesis $H_O$ that the means of groups by *gender* are similar, means of groups by *diet.type* are similar and there is no significant interaction between *gender* and *diet.type* features.

\newpage

## Post-Hoc Analysis

```{r}
#post-hoc analysis
TukeyHSD(model)
```

Having rejected the null hypothesis $H_O$. We wanted to know which factor plays the biggest role in Weight Loss.

\newpage

## Conclusion

After conducting a through pre-testing and then a Two-Way Analysis, we can safely say that Diet Type plays the most important role in Weight Loss. Even within the Various Diet Types the best Type for Weight Loss is that of Diet 'C'.

This Analysis should allow anyone that followed it so far to understand what potential implications it could have at an industrial level.

### Citations and References

Couturier, D. L., Nicholls, R., and Fernandes, M. (2018). ANOVA with R: analysis of the diet data set. Retrieved online.
